Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling

### 2022-08-31 manuscript/manuscript.tex
Move main.py out of bin/

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{authblk}
\graphicspath{ {./images/} }
\linespread{1.25}
\textwidth=380pt

\title{Sequence modeling enables direct inference of germline haplotypes from short read sequencing data}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning that directly infers germline haplotypes from short read pileups. Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

More recent tools have incorporated elements of deep learning into the allele likelihood calculation or allele generation steps. DeepVariant (Poplin et al. 2018) uses a statistical method based on the HaplotypeCaller approach to identify candidates, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections. HELLO (Ramachandran et al. 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input. Clair (Luo et al. 2019, Luo et al. 2020), uses a novel multitask deep learning approach to predict several properties of a potential variant at a given site, including zygosity and allele length.

@@

Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position across reads, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions. The 2D encoding implementation follows Wang \& Jyh-Charn (2021).

The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$. The full model architecture is shown in \ref{fig:architecture}.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./jovian_architecture_diagram.png}
	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors  }
	\label{fig:architecture}
\end{figure}

Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format. For each region, all reads overlapping the window were obtained from the corresponding alignment file. Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.  Reads are then sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features are encoded; the first four are the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No gap tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features. Positions in the input not corresponding to an aligned base were all set to zero.

@@

Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence. Two sequences were generated for each region, one representing each haplotype. In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status. Briefly, all possible genotypes (pairs of haplotypes) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing. This phasing procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.

\begin{center}
	\begin{tabular}{ lc }
	 \textbf{Region type} & \textbf{Number of regions (per sample)} \\
	 No variant & 200,000 \\
	 SNV & 200,000 \\
	 Insertions & 200,000 \\
	 Deletions & 200,000 \\
	 Insertion/Deletion + SNV & 50,000 \\
	 Insertion + Deletion & 30,000 \\
	 SNV (low complexity) & 80,000 \\
	 Insertion (low complexity) & 80,000 \\
	 Deletion (low complexity) & 80,000 \\  
	 Insertion + Deletion (low complexity) & 20,000 \\
	 Insertion/Deletion + SNV (low complexity) & 80,000 \\
	\end{tabular}
	\label{table:regioncounts}
\end{center}

@@

Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e^{-5}$ and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).

For a given haplotype we use a modified cross-entropy loss function to compare the predictions to the target sequence. However, two haplotypes exist for each region and they have no inherent order. For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random. To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.

@@

The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb. True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al. 2015). We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and `balanced' class weighting option.

\begin{center}
	\begin{tabular}{ l }
	 \textbf{Feature} \\
	 Mean probability of base call at all positions affected by variant \\ 
	 Number of reference bases affected \\
	 Number of alternate bases affected \\
	 Minimum probability of base call at all positions affected by variant \\
	 Maximum probability of base call at all positions affected by variant \\
	 Number of variants in \textit{cis} (same haplotype) with this variant \\
	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
	 Total number of variants in both \textit{cis} and \textit{trans} \\
	 Distance in bases of call from beginning of window - minimum across windows \\
	 Distance in bases of call from beginning of window - maximum across windows \\
	 Read depth at variant position \\
	 Gnomad genome frequency (v2.0.2) of variant \\
	 Zygosity (heterozygous or homozygous) of variant \\
	 Fraction of reads in BAM/CRAM supporting variant call \\
	 Fraction of reads with mapping quality > 10 supporting variant call \\
	 Total number of reads supporting variant call \\
	\end{tabular}
	\label{table:features}
\end{center}

@@

\begin{center}
	\begin{tabular}{ lllll }
	 Model name & Encoder layers & Attention heads & Hidden dimension & Embedding dimension \\ 
	\hline
	 5m & 4 & 4 & 200 & 600 \\ 
	 10m & 6 & 6 & 250 & 400 \\ 
	 25m & 8 & 8 & 400 & 800 \\ 
	 \hline
	\end{tabular}
\end{center}

During training, sensitivity and precision as measured on the holdout regions from chromosomes 21 and 22 increase for all variant classes, with returns dwindling after 30-40 epochs of training. Little difference was observed between the "full" and "half" sized training data sets, but both sensitivity and precision were lower for the quarter-sied training set. These results indicate that additional training examples beyond the "half" size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.

@@

@@

To characterize variant detection accuracy, we called variants on validation regions from chromosomes 21 and 22 for each of the 17 samples, and compared results to DeepVariant (v1.3.0, Poplin et al. 2018), HaplotypeCaller (v. 4.1.4.1), and Strelka2 (v. 2.9.10, Kim et al. 2018). For this comparison we used 25M model trained for 25 epochs on the 'huge' training set (~$20^6$ training examples). We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples. Jovian calls were filtered at 0.25, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0. Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al. 2019) using Genome-in-a-Bottle (v.4.2.1) as the benchmark variant set.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./sensitivity-precision.png}
	\caption{ Sensitivity and precision of Jovian variant calls (10M parameter model, large training size) compared to other callers }
	\label{fig:sens-ppv}
\end{figure}

Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2 (\ref{fig:sens-ppv}). For SNV calls Jovian has the highest sensitivity (99.1\%) and precision similar to HaplotypeCaller (99.5\%), but precision is lower than DeepVariant and Strelka2 (99.8\% and 99.9\%, respectively).

Across indel size classes Jovian has high sensitivity for deletions less than about 15bp, but relatively low sensitivity for larger deletions (~75\%). For insertions Jovian has sensitivity similar to other callers at smaller sizes, but again reduced sensitivity for the largest class (~68\%). Precision follows a similar pattern, but with relatively high precision for smaller indel sizes and reduced precision at larger sizes.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./accuracy_indel_size.png}
	\caption{ Sensitivity and precision of Jovian indel calls by size class (10M parameter model, large training size) compared to other callers }
	\label{fig:indelsize-sens-ppv}
\end{figure}

@@

To check the accuracy of predicted genotypes and phasing of variants we compared calls in the chromosome 21 and 22 evaluation regions to the known genotype and phase from Genome-in-a-Bottle (GIAB) variant data. Approximately 99.5\% of homozygous and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both genotypes accurately.  For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB. The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so analysis was restricted to those samples. Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs.

@@

To identify variants requiring complex reconstruction, we first obtained all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22. For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp soft-clipped as requiring 'complex' reconstruction. These variants are moved to a separate truth set. After running our variant calling procedure on the alignment file, we compute the fraction of variants from the truth set that were identified by our model. This procedure was repeated for all 17 samples in our data set.

Our model was able to reconstruct the correct haplotypes for ~89.5\% (mid 50\% 82.5-94.8) of these variants, somewhat lower than other callers (\ref{table:complex-sensitivity}).  These results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads. Instead, the model learns to synthesize information across tokens to produce complete haplotypes with accuracy similar to (but slightly less than) the top performing current methods.

\begin{center}
	\begin{tabular}{ ccc }
	 Caller & Sensitivity & Mid 50\%  \\ 
	\hline
	 Jovian(10M) & 89.5\% & 82.5 - 94.8 \\ 
	 HaplotypeCaller & 95.8\% & 91.0 - 100.0  \\ 
	 DeepVariant & 91.7\% & 83.5 - 97.1  \\ 
	 Strelka & 93.1\% & 90.9 - 98.4 \\
	 \hline
	\end{tabular}
	\label{table:complex-sensitivity}
\end{center}

\subsection{Use of local context}

We were interested to see how model variant calling was informed by contextual information from nearby alignment columns. In particular, the model must learn to discriminate real variation from noise at low Variant Allele Frequency (VAF, the fraction of reads supporting a given allele candidate), and this model decision process may involve the use of local context information. For example, the model may be less likely to call a SNV in a region containing many reference mismatches, because the presence of multiple mismatches suggests poor sequencing quality or other technical artifacts. However, the call probability is likely to be modified by other factors, including VAF and the base call quality associated with the prospective variant.

To examine the influence of the number of nearby mismatches, VAF, and base quality on the likelihood that a prospective SNV was called, we first collected genomic positions where 10 - 30\% of reads supported a non-reference basecall. We then tallied the number of genomic positions at which 10\% - 20\% of reads supported a non-reference call in a window 20bp upstream and downstream of the candidate. We also stratified the candidate position into low base quality and high base quality groups, and smaller VAF windows of 10-20\%, 15-25\% and 20-30\%.

\begin{figure}[htp]
	\includegraphics[scale=0.25]{./variant_context_BQ.png}
	\caption{ Effect of number of nearby reference mismatches on the likelihood that a poorly supported candidate variant will be called. }
	\label{fig:variant_context}
\end{figure}

The probability that a prospective variant is called is strongly associated with the number of nearby alignment columns containing mismatches (Figure \ref{fig:variant_context}). For instance, prospective variants with VAF between 20-30\% and high base quality were called in over 60\% of examined cases when there were no nearby mismatches, but called in less than 30\% of cases when a single other column contains a mismatch, and in less than 10\% of cases when 5 or more nearby positions contain a mismatch. A similar, but less pronounced trend exists when the prospective variant is at lower VAF (left two panels, figure \ref{fig:variant_context}). When the prospective variant has poor base quality scores, the association disappears and the likelihood that the variant is called remains uniformly low. Together, these results indicate that the models learns a nuanced SNV detection procedure that incorporates multiple elements, including base quality, VAF, and information from local surrounding bases to discriminate actual genetic variation from technical artifacts and sequencing errors.

@@

To assess the impact of the two-dimensional positional encoding, we replaced the 2D encoding module with traditional 1D encoding, and left other model parameters unmodified. We conducted two training runs, one with the 10M model and smaller training data size, and the other with the 25M model and medium training data size. Both training runs were conducted for 25 epochs. When compared to their equivalent models with 2D encoding, the 1D encoding model showed substantially reduced precision, in particular for the SNV class (Figure \ref{fig:encoding_ablation}). Sensitivity was also reduced but to a lesser degree.

Two-dimensional encoding allows the model to track which read contains a given base across genomic positions. We hypothesize that read-level tracking is important for precision because mismapped or poorly aligned reads generate many reference mismatches, most of which are likely to be errors. The model must learn to identify and ignore such reads, and 2D encoding facilitates this because the model can more easily track read-level features across input tokens.

@@

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity near 90\%. In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~97-99\% of cases where variants are within 100bp.  The model also learns to consider the surrounding read context when predicting the presence of a SNV. For instance, a potential SNV in a region with few reads containing reference mismatches is much more likely to be called than a similar variant in a region where many reads contain mismatches, indicating that the model has learned to be cautious of potential SNVs in `dirty' regions where many reads contain errors or are misaligned.

Overall variant detection accuracy is similar to other modern callers. At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (\ref{fig:sens-ppv}). The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.6\% for deletions, 68.8\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.0\%).  For SNVs Jovian demonstrated the highest sensitivity among all callers (mean 99.1\%), and precision similar to HaplotypeCaller but lower than DeepVariant and Strelka2 (Jovian mean precision 99.5\%).

@@

 Source code for is available via git at https://github.com/ARUP-NGS/somewhere

>>>>

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{authblk}
\graphicspath{ {./images/} }
\linespread{1.25}
\textwidth=400pt

\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups. Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

More recent tools have incorporated elements of deep learning into the allele likelihood calculation or allele generation steps. DeepVariant (Poplin et al. 2018) uses a statistical method based on the HaplotypeCaller approach to identify candidates, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections. HELLO (Ramachandran et al. 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and position dimensions of the input. Clair (Luo et al. 2019, Luo et al. 2020), uses a novel multitask deep learning approach to predict several properties of a potential variant at a given site, including zygosity and allele length.

@@

Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding, we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions. The 2D encoding implementation follows Wang \& Jyh-Charn (2021).

The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully-connected components that each generate a single haplotype. The decoder is not autoregressive and does not utilize special start-of-sequence or end-of-sequence tokens. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$. The full model architecture is shown in Figure \ref{fig:architecture}.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./jovian_architecture_diagram.png}
	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors}
	\label{fig:architecture}
\end{figure}

Input tensors were generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format. For each region, all reads overlapping the window were obtained from the corresponding alignment file. Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.  Reads were sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features are encoded; the first four are the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No gap tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features. Positions in the input not corresponding to an aligned base were all set to zero.

@@

Target haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence. Two sequences were generated for each region, one representing each haplotype. In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status. Briefly, all possible genotypes (pairs of haplotypes) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing. This phasing procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.

\begin{center}
	\begin{table}
	\begin{tabular}{ lc }
	 \textbf{Region type} & \textbf{Number of regions (per sample)} \\
	 No variant & 200,000 \\
	 SNV & 200,000 \\
	 Insertions & 200,000 \\
	 Deletions & 200,000 \\
	 Insertion/Deletion + SNV & 50,000 \\
	 Insertion + Deletion & 30,000 \\
	 SNV (low complexity) & 80,000 \\
	 Insertion (low complexity) & 80,000 \\
	 Deletion (low complexity) & 80,000 \\  
	 Insertion + Deletion (low complexity) & 20,000 \\
	 Insertion/Deletion + SNV (low complexity) & 80,000 \\
	\end{tabular}
	\caption{Number of regions selected from each sample to generate training data}
	\label{table:regioncounts}
\end{table}
\end{center}

@@

Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).

For a given haplotype we use a modified cross-entropy loss function to compare the predictions to the target sequence. For each region two haplotypes exist and they have no inherent order. For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn to predict which because the assigment is random. To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.

@@

The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb. True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al. 2015). We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and `balanced' class weighting option. Features are listed in Table \ref{table:features}.

\begin{center}
	\begin{table}
	\begin{tabular}{ l }
	 \textbf{Feature} \\
	 Mean probability of base call at all positions affected by variant \\ 
	 Number of reference bases affected \\
	 Number of alternate bases affected \\
	 Minimum probability of base call at all positions affected by variant \\
	 Maximum probability of base call at all positions affected by variant \\
	 Number of variants in \textit{cis} (same haplotype) with this variant \\
	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
	 Total number of variants in both \textit{cis} and \textit{trans} \\
	 Distance in bases of call from beginning of window - minimum across windows \\
	 Distance in bases of call from beginning of window - maximum across windows \\
	 Read depth at variant position \\
	 Gnomad genome frequency (v2.0.2) of variant \\
	 Zygosity (heterozygous or homozygous) of variant \\
	 Fraction of reads in BAM/CRAM supporting variant call \\
	 Fraction of reads with mapping quality > 10 supporting variant call \\
	 Total number of reads supporting variant call \\
	\end{tabular}
	\label{table:features}
	\caption{ Features used to train classifier model for estimation of variant quality scores }
\end{table}
\end{center}

@@

\begin{center}
	\begin{table}
	\begin{tabular}{ ccccc }
	 Model name & Encoder layers & Attn. heads & Hidden dim. & Embedding dim. \\ 
	\hline
	 5m & 4 & 4 & 200 & 600 \\ 
	 10m & 6 & 6 & 250 & 400 \\ 
	 25m & 8 & 8 & 400 & 800 \\ 
	 \hline
	\end{tabular}
	\caption{ Model parameter combinations used for the three model sizes investigated }
\end{table}
\end{center}

@@

During training, sensitivity and precision increased for all variant classes, with returns dwindling after 30-40 epochs of training (Figure \ref{fig:trainsize}). Little difference was observed between the Full and Half sized training data sets, but both sensitivity and precision were lower for the Quarter-sized training set. These results indicate that additional training examples beyond the Half size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./modelsize_effect.png}
	\caption{ Sensitivity and precision curves for the Half-sized data set, for the 25M, 10M, and 5M model sizes. Dashed lines are raw values, solid lines indicate rolling average with window size of 5. }
	\label{fig:modelsize}
\end{figure}

Sensitivity was similar across all model sizes evaluated for all variant classes, but precision increased as model size was increased from 5M to 10M (Figure \ref{fig:modelsize}). However, precision did not increase further as model size was increased from 10M to 25M. Together with the training data size analysis, the results indicate that neither increasing model capacity beyond 10M parameters or training data size beyond approximately 12M examples results in improved model performance.

@@

To characterize variant detection accuracy, we called variants on validation regions from chromosomes 21 and 22 for each of the 17 samples, and compared results to DeepVariant (v1.3.0, Poplin et al. 2018), HaplotypeCaller (v. 4.1.4.1), and Strelka2 (v. 2.9.10, Kim et al. 2018). For this comparison we used the 10M model trained on the Full set. We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples. Jovian calls were filtered at quality 0.25, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0. Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al. 2019) using Genome-in-a-Bottle (v4.2.1) as the benchmark variant set.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./sensitivity-precision.png}
	\caption{ Sensitivity and precision of Jovian variant calls (10M parameter model, Full training size) compared to other callers }
	\label{fig:sens-ppv}
\end{figure}

Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2 (Figure \ref{fig:sens-ppv}). For SNV calls Jovian has the highest sensitivity (99.1\%) and precision similar to HaplotypeCaller (99.5\%), but precision is lower than DeepVariant and Strelka2 (99.8\% and 99.9\%, respectively).

Across indel size classes Jovian has high sensitivity for deletions less than about 15bp, but relatively low sensitivity for larger deletions ($\approx$75\%). For insertions Jovian has sensitivity similar to other callers at smaller sizes, but again reduced sensitivity for the largest class ($\approx$68\%). Precision follows a similar pattern, but with relatively high precision for smaller indel sizes and reduced precision at larger sizes.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./accuracy_indel_size.png}
	\caption{ Sensitivity and precision of Jovian indel calls by size class (10M parameter model, Full training size) compared to other callers }
	\label{fig:indelsize-sens-ppv}
\end{figure}

@@

To check the accuracy of predicted genotypes and phasing of variants we compared calls in the chromosome 21 and 22 evaluation regions to the known genotype and phase from Genome-in-a-Bottle (GIAB) variant data. Approximately 99.5\% of homozygous and 99.0\% of heterozygous calls matched the true genotype. For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB. The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so analysis was restricted to those samples. Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs. Together, these analyses indicate that the model learns to reconstruct variant phase and genotype accurately, producing pairs of haplotype sequences that match those present in the sample with high fidelity.

@@

To identify variants requiring complex reconstruction, we first obtained all indel variants from Genome-in-a-Bottle greater than 4 bp in length from the validation regions on chromosomes 21 and 22. For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp soft-clipped as requiring `complex' reconstruction. These variants were moved to a separate truth set. After running our variant calling procedure on the alignment file, we computed the fraction of variants from the truth set that were identified by our model. This procedure was repeated for all 17 samples in our data set.

Our model was able to reconstruct the correct haplotypes for $\approx$89.5\% (mid 50\% 82.5-94.8) of these variants, somewhat lower than other callers (table \ref{table:complex-sensitivity}).  These results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads. Instead, the model learns to synthesize information across tokens to produce complete haplotypes with accuracy similar to (but slightly less than) the top performing current methods.

\begin{center}
\begin{table}	 
	\begin{tabular}{ ccc }
	 Caller & Sensitivity & Mid 50\%  \\ 
	\hline
	 Jovian(10M) & 89.5\% & 82.5 - 94.8 \\ 
	 HaplotypeCaller & 95.8\% & 91.0 - 100.0  \\ 
	 DeepVariant & 91.7\% & 83.5 - 97.1  \\ 
	 Strelka & 93.1\% & 90.9 - 98.4 \\
	 \hline

	\end{tabular}
	\label{table:complex-sensitivity}
	\caption{ Sensitivity of detection algorithms for variants requiring complex reconstruction }
\end{table}
\end{center}

\subsection{Use of local context for SNV prediction}

Variant callers must integrate multiple types of information to produce accurate calls, including the fraction of reads supporting a given call (Variant Allele Frequency, VAF),base and mapping qualities, read position, and potentially information from nearby positions. To examine the influence of the number of nearby reference mismatches, VAF, and base quality on the likelihood that a prospective SNV was called, we first collected genomic positions in which 10 - 30\% of reads supported a non-reference basecall. We then tallied the number of genomic positions at which 10\% - 20\% of reads supported a non-reference call in a window 20bp upstream and downstream of the candidate. We also stratified the candidate position into low base quality and high base quality groups, and smaller VAF windows of 10-20\%, 15-25\% and 20-30\%.

\begin{figure}[htp]
	\includegraphics[scale=0.22]{./variant_context_BQ.png}
	\caption{ Effect of number of nearby reference mismatches on the likelihood that a poorly supported candidate variant will be called. }
	\label{fig:variant_context}
\end{figure}

The probability that a prospective variant is called is strongly associated with the number of nearby alignment columns containing mismatches (Figure \ref{fig:variant_context}). For instance, prospective variants with VAF between 20-30\% and high base quality were called in over 60\% of examined cases when there were no nearby mismatches, but called in less than 30\% of cases when a single other column contains a mismatch, and in less than 10\% of cases when 5 or more nearby positions contain a mismatch. A similar, but less pronounced trend exists when the prospective variant is at lower VAF (left two panels, figure \ref{fig:variant_context}). When the prospective variant has poor base quality scores, the association disappears and the likelihood that the variant is called remains uniformly low. Overall, these results indicate that the model learns a nuanced SNV detection procedure that discards most prospective SNVs with low base quality, but selectively retains those with higher base quality provided that not too many nearby positions contain reference mismatches.

@@

To assess the impact of the two-dimensional positional encoding, we replaced the 2D encoding module with traditional 1D encoding, and left other model parameters unmodified. We conducted two training runs, one with the 10M model and Quarter training data size, and the other with the 25M model and Half training data size. Both training runs were conducted for 25 epochs. When compared to their equivalent models with 2D encoding, the 1D encoding model showed substantially reduced precision, in particular for the SNV class (Figure \ref{fig:encoding_ablation}). Sensitivity was also reduced but to a lesser degree.

By applying positional encoding across both the read and position (input token) dimensions, two-dimensional encoding allows the model to track which read contains a given base across positions. We hypothesize that read-level tracking is important for precision because mismapped or poorly aligned reads generate many reference mismatches, most of which are likely to be errors. The model must learn to identify and ignore such reads, and 2D encoding facilitates this because the model can more easily track read-level features across input tokens.

@@

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity near 90\%. In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in 97-99\% of cases where variants are within 100bp.  The model also learns to consider the surrounding read context when predicting the presence of a SNV. For instance, a potential SNV in a region with few reads containing reference mismatches is much more likely to be called than a similar variant in a region where many reads contain mismatches, indicating that the model has learned to be cautious of potential SNVs in `dirty' regions where many reads contain errors or are misaligned.

Overall variant detection accuracy is similar to other modern callers. At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (Figure \ref{fig:sens-ppv}). The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.6\% for deletions, 68.8\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.0\%).  For SNVs Jovian demonstrated the highest sensitivity among all callers (mean 99.1\%), and precision similar to HaplotypeCaller but lower than DeepVariant and Strelka2 (Jovian mean precision 99.5\%).

@@

 Source code for is available via git at https://github.com/brendanofallon/jovian
